Self Referencing Sequences

### 2014-11-07 draft.tex
Edits to Iterative Process

@@

One possible iterative process is as follows. Let the sequence $s^0$ be 2, and let the sequence $s^{(i)}$ for $i \geq 1$ be the unique sequence that starts with 2 and is such that its block lengths, read from left to right, reproduce the sequence $s^{(i-1)}$. This gives:
\begin{align*}
s^{(0)}&= 2\\
s^{(1)} &= 22\\
s^{(2)} &= 2211\\
s^{(3)}&= 221121\\
s^{(4)} &= 221121221\\
&\vdots
\end{align*}
Let $s^\infty$ be the limit of this process. 
By limit, we mean the limit with respect to each term seperately, so we can define
\[ s^{(i)}_j \rightarrow s^\infty_j \]
as $i \rightarrow \infty$. We claim that $s^\infty$ is the unique self-referential sequence $s(C,2)$. In fact we can prove a more general result about this iterative process.

\begin{theorem}
Let $C$ be some generating cycle. Let $s^0 = a$ for some $a \in C$ where $a \neq 1$, and let $s^{(i)}$ be the sequences generated by the iterative process. Then the limit of this process $s^\infty$ exists and is the sequence $s(C,a)$.
\end{theorem}

@@

For the inductive step, we observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that the sequence it describes is longer because the sequence it describes has length equal to the sum of the original prefix. In other words, the length of $s^{(i+1)}$ equals the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$ (since there are numbers that are greater than 1).

The resultant sequence is a prefix of the self referencing sequence because if we apply our process to any prefix of the self referencing sequence it will produce a prefix of the self referencing sequence.  
\end{proof}

Note that we must specify that $a \neq 1$, because that is the only case in which the length of $s^{(i+1)}$ is not strictly greater than the length of $s^{(i)}$.

It turns out that we can be less restrictive about what our starting sequence, $s^{(0)}$, is. Going back to the case of $C = (1 2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2. And still, the limiting sequence will be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(C,2)$. We note that $s^\infty = s(C,a)$ because because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(C,a)$ in the limit. We call the rest of each $s^{(i)}$ that is not a prefix of $s(C,2)$ the \emph{tail}.

We now consider two different forms of convergence. We know that $s^{(i)}$ in some sense approaches $s(C,a)$, as arbitrarily many elements at the beginning of the sequence will match up with the beginning of $s(C,a)$ as $i$ increases. However, if the tail of the sequence never entirely disappears, then there does not exist an $i$ where $s^{(i)}$ is exactly a prefix of $s(C,a)$. Thus, we formalize two types of convergence
\begin{definition} We say that a sequence $s^{i}$ is weakly convergent to $s(C,a)$ if $s^\infty = s(C,a)$. We say that the sequence $s^{i}$ is strongly convergent if $s^{(i)}$ is a prefix of $s(C,a)$ for all $i \geq k$ for some $k$.
\end{definition}
It is unclear what conditions must be placed on the starting sequence in order to ensure strong convergence.%, although based on numerical simulation we make the following conjecture. 
%
%\begin{conj} Let $C = (1,2)$. A sequence $s^{(i)}$ is weakly convergent but not strongly convergent to $s(C, a)$ if the starting sequence begins with an $a \in \{1,2\}$ and contains 4 or more elements.
%\end{conj} 
%We make this conjecture on the basis on computer simulations which executes the iterative process to at least 20 iterations. More iterations were not completed because after this point the simulations began to take impractically long amounts of time to complete.

@@

\section{Density Bounds for $C = (12)$}
\label{densitybounds}

@@

\begin{itemize}
\item In Section~\ref{equivalence}, we propose the idea that in some sense, the number that we start a sequence with does not matter. In other words, any sequence that uses a certain generating cycle is equivalent to any other sequence created by the same generating cycle.
%\item In the same section, we propose a condition for determining when two sequences are not equivalent. We conjecture that if two sequences have unequal numbers with the same read distances, then they are not equivalent. It seems from trying cases that this condition is sufficient, but we have been unable to prove it.
\item In Section~\ref{iterativeprocess} we explored when sequences are strongly and weakly convergent. We have not yet been able to investigate strong and weak convergence in great detail, as we have been unable to make precise mathematical statements about these types of convergence. Most of what we know about them comes from analyzing numerical data.
\item In Section~\ref{substitutionrules}, we look at substitution rules for certain ``nice" generating cycles of numbers, such as $(13)$ and $(24)$. These sets allow for substitution rules because they satisfy the trio of properties we outline in that section. However, we do not know that these are the only properties. We leave unresolved the question of determining precisely which sets allow for substitution rules and which ones ones do not. The major difficulty here seems to come in proving that a certain generating cycle does \emph{not} allow for substitution rules.
\item In Section~\ref{density} we were able to compute the density of any sequence with generating cycle $(13)$, and we were able to give bounds on the density of any sequence with generating cycle $(12)$. However, we leave unresolved the question of what the density for $(12)$ actually is, or even if we can obtain tighter bounds.
\end{itemize}

>>>>

\emph{Written by Fermi Ma and Perry Kleinhenz edited by Erik Waingarten}

@@

One possible iterative process is as follows. Let the sequence $s^0$ be 2, and let the sequence $s^{(i)}$ for $i \geq 1$ be the unique sequence that starts with 2 and is such that its block lengths, read from left to right, reproduce the sequence $s^{(i-1)}$. This gives:
\begin{align*}
s^{(0)}&= 2\\
s^{(1)} &= 22\\
s^{(2)} &= 2211\\
s^{(3)}&= 221121\\
s^{(4)} &= 221121221\\
&\vdots
\end{align*}
We define the limit of this process, the sequence $s^{\infty}$ as 
\begin{equation*}
s_{j}^{\infty} = \lim_{i \rightarrow \infty} s_{j}^{(i)}.
\end{equation*}
We claim that this $s^{\infty}$ exists and is the unique self-referential sequence $s(C,2)$. In fact we can prove a more general result about this iterative process.

Before we do so however we will clarify two different notions of convergence, that will allow us to state our result more precisely. 
\begin{definition} We say that a sequence $s^{(i)}$ is weakly convergent to $s(C,a)$ if $s^{\infty}$ exists and $s^{\infty} = s(C,a)$. We say that the sequence $s^{(i)}$ is strongly convergent to $s(C,a)$ if $s^{(i)}$ is a prefix of $s(C,a)$ for all $i \geq k$ for some $k\geq 0$. 
\end{definition}
Note that all strongly convergent sequences are also weakly convergent.

\begin{theorem}
Let $C$ be some generating cycle. Let $s^0 = a$ for some $a \in C$ where $a \neq 1$, and let $s^{(i)}$ be the sequences generated by the iterative process. Then $s^{(i})$ is strongly convergent to $s(C,a)$.
\end{theorem}

@@

For the inductive step, we observe that a prefix of the sequence describes the block lengths of a \emph{longer} prefix of the sequence. We know that the sequence it describes is longer because the length of $s^{(i+1)}$ equals the sum of of the numbers in $s^{(i)}$, which is strictly greater than the length of $s^{(i)}$ (since there are numbers that are greater than 1).

%The resultant sequence is a prefix of the self referencing sequence because if we apply our process to any prefix of the self referencing sequence it will produce a prefix of the self referencing sequence.

Since very $s^{(i)}$ is a prefix of $s(C,a)$, we know that once $s^{(i)}_{j}$ has been assigned a value, $s^{(k)}_{j}$ will have the same value for all $k \geq i$. Because of this $\lim_{i\rightarrow \infty} s^{(i)}_{j}$ is well defined and so $s^{\infty}$ exists. As the limit of the process we know that $s^{\infty}$ is a prefix of $s(C,a)$, but since at every step of the process the length of the sequence strictly increases we must have that $s^{\infty}$ has infinite length, and so $s^{\infty} = s(C,a)$.

\end{proof}

Note that we must specify that $a \neq 1$, otherwise $s^{(i)}$ is always $1$, and so the length of the sequence does not grow.

Above we only considered starting sequences which have length one,  but we can be less restrictive about what our starting sequence, $s^{(0)}$, is. Going back to the case of $C = (1 2)$, we can let $s^{(0)}$ be \emph{any} sequence of 1's and 2's that starts with a 2 and the limiting sequence will still be the self-referential sequence.
For example, suppose $s^{(0)} = 2,1,1,1,2$. The rules for generating $s^{(i)}$ give:
\begin{align*}
s^{(0)}&= \mathbf{2},1,1,1,2\\
s^{(1)} &= \mathbf{2,2,1},2,1,2,2\\
s^{(2)} &= \mathbf{2,2,1,1,2,1},1,2,1,1,2,2\\
s^{(3)} &= \mathbf{2,2,1,1,2,1,2,2,1,2},1,1,2,1,2,2,1,1\\
&\vdots
\end{align*}
Here, we bold the numbers that form a prefix of $s(C,2)$. We note that $s^\infty = s(C,a)$ because because $s^{(0)}$ contains a 2 at the beginning, which by the above theorem will produce $s(C,a)$ in the limit. This is true in general.

\begin{corollary}
Let $C$ be some generating cycle, let $s^(0)$ be any sequence of elements of $C$ which begins with some $a \in C$ such that $a \neq 1$. Then $s^{(i)}$ is weakly convergent to $s(C,a)$. 
\end{corollary}
\begin{proof} 
\end{proof}

It is unclear what conditions must be placed on the starting sequence in order to ensure strong convergence, although based on numerical simulation we suspect that applying the process to the starting sequence $2222$ 
%
%\begin{conj} Let $C = (1,2)$. A sequence $s^{(i)}$ is weakly convergent but not strongly convergent to $s(C, a)$ if the starting sequence begins with an $a \in \{1,2\}$ and contains 4 or more elements.
%\end{conj} 
%We make this conjecture on the basis on computer simulations which executes the iterative process to at least 20 iterations. More iterations were not completed because after this point the simulations began to take impractically long amounts of time to complete.

@@

\section{Density Bounds for $C = (12)$}
\label{densitybounds}
\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz}

@@

\begin{itemize}
\item In Section~\ref{equivalence}, we propose the idea that in some sense, the number that we start a sequence with does not matter. In other words, any sequence that uses a certain generating cycle is equivalent to any other sequence created by the same generating cycle.
%\item In the same section, we propose a condition for determining when two sequences are not equivalent. We conjecture that if two sequences have unequal numbers with the same read distances, then they are not equivalent. It seems from trying cases that this condition is sufficient, but we have been unable to prove it.
\item In Section~\ref{iterativeprocess} we explored when sequences are strongly and weakly convergent. We have not yet been able to investigate strong and weak convergence in great detail, as we have been unable to make precise mathematical statements about these types of convergence. Most of what we know about them comes from analyzing numerical data.
\item In Section~\ref{substitutionrules}, we look at substitution rules for certain ``nice" generating cycles of numbers, such as $(13)$ and $(24)$. These sets allow for substitution rules because they satisfy the trio of properties we outline in that section. However, we do not know that these are the only properties. We leave unresolved the question of determining precisely which sets allow for substitution rules and which ones ones do not. The major difficulty here seems to come in proving that a certain generating cycle does \emph{not} allow for substitution rules.
\item In Section~\ref{densitycomputation} we were able to compute the density of any sequence with generating cycle $(13)$, and in Section ~\ref{densitybounds} were able to give bounds on the density of any sequence with generating cycle $(12)$. However, we leave unresolved the question of what the density for $(12)$ actually is, or if we can obtain tighter bounds.
\end{itemize}
